---
title: "Delaware Bay Test Summary"
format: docx
editor: visual
---

```{r, echo = FALSE, warning=FALSE, message=FALSE}

library(fs)
library(dplyr)
library(sf)
library(lubridate)
library("rnaturalearth")
library("rnaturalearthdata")
library(fpc)
library(dbscan)
library(ggplot2)
library(adehabitatHR)
library(sp)
library(viridisLite)
library(ggpointdensity)
library(ggspatial) # For base maps

```

To better understand Red Knot migration and stopover locations we used high precision GPS data locations from 300 individual GPS/satellite tags to identify and characterize stopovers and migration routes.

<br>

### 1.GPS/satellite data

Location and deployment data was obtained from 300 individual Red Knot *rufa* *subspecies*. This data was reviewed to remove spatial outliers and erroneous locations, such as those well beyond the flight path of the individual. Data filtering and QA was completed using [R coding](https://github.com/ninoxconsulting/REKN_gps) and QGIS (version 3.28.11).

### 2. Methodology

To identify stopover locations, we expanded on initial testing conducted by Theo Diehl and Stephanie Feigin. The aim was to standardize and refine the process to map stopover locations, to enable transparent and repeatable method to be rolled out to multiple individuals.

#### 2.1 GPS data pre processing

All GPS/satellite data was downloaded in .csv format along with detailed deployment data. This was was previously compiled by G.Perkins as part of ongoing work with US F&W.

Data was filtered to include individuals from rufa subpopulations who had recorded a movement away from their initial deployment location. This reduced the error of inclusion of short-lived tags, tags that were malfunctioning or tags that had either dropped off their birds or the bird had died.

Location data was converted from text to a spatial format (using the sf package) and transformed to a equidistant global projection [(ESPG: 4087)](https://spatialreference.org/ref/epsg/4087/). This projection enables accurate distance calculations as it is an equal area projection, hence distance measurements can be calculated and compared at any point on the globe.

#### 2.2 Identify Stopover Locations

We applied a density based clustering algorithm (DBScan) to identify clusters of locations or stopovers ( Ester et al. 1996). The advantage of this method over more traditional clustering (i.e. K-means clustering), is that the number of clusters does not need to be defined apriori, clusters can be non-circular, and outliers are identified. This is well suited to Red Knot GPS/satellite data which can have multiple stopovers in close proximity and where outliers are expected due to variable spatial precision.

In the example below we show how the dbscan is used to identify clustered locations for one tag (id = 260808) within Delaware Bay. This bird was tagged in Avalon in October 2024.

As a test we will focus only on Delaware Bay and revise methodology as we expand to broader stopover locations.

```{r, echo = FALSE, warning=FALSE,fig.align='center'}
dbi <- read.csv(path("01_raw_data", "test_db_locations.csv"))

data <- dbi |> 
  select(-tag.id)

dbs2500 <- dbscan(data, 2500, MinPts = 5)
dbs5000 <- dbscan(data, 5000, MinPts = 5)
#method = c("hybrid", "raw", "dist")

## Remove the points that are outside the main clusters 
## add the cluster number back to full dataset and filter out
db1_all <- dbi |> 
  mutate(cluster_2500 = dbs2500$cluster) |> 
  mutate(eps = 2500)

db1_all1 <- dbi |>  
   mutate(cluster_2500 = dbs5000$cluster) |> 
    mutate(eps = 5000)

dball <- bind_rows(db1_all, db1_all1)

dbisf<- st_as_sf(dball, coords = c("X", "Y"), crs = 4087)
  
#dbisf <- dbisf |> 
#  filter(cluster !=0)

#plot(dbisf)

# Geographic distributon of tags
world <- ne_countries(scale = "medium", returnclass = "sf")

Americas <- world %>% 
  dplyr::filter(region_un == "Americas")%>% 
  select(admin)

# entire north America 
global <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  #geom_sf(data = all_seasons, aes(colour = id), alpha = 0.1, colour = "dark blue") +
  geom_sf(data = dbisf, aes(colour = cluster_2500), size = 1.2, alpha = 0.6) +
  #facet_wrap(~movement_dir)+
  # geom_point(ru, aes(x = lng, y = lat), size = 4) +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
  #coord_sf(xlim = c(-130, -40), ylim = c(10, 80), expand = FALSE)+
  scale_colour_viridis_c() +
  #coord_sf() +
  facet_wrap(vars(eps)) +
  theme_bw()+
  labs(title = paste0("Red Knot distance threshold (tag = 260808)"))+
  guides(col= guide_legend(title= "Clusters"))+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

#global

```

We can review the locations for this individual in the Delaware Bay region (Figure 1) and calculate the distance between each points.

```{r, echo = FALSE, fig.align='center', fig.cap="Figure 1: GPS locations for tag 260808 within Delaware Bay", fig.width=5, fig.height=5}

# Geographic distributon of tags
world <- ne_countries(scale = "medium", returnclass = "sf")

Americas <- world %>% 
  dplyr::filter(region_un == "Americas")%>% 
  select(admin)

# entire north America 
global_basic <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  #geom_sf(data = all_seasons, aes(colour = id), alpha = 0.1, colour = "dark blue") +
  geom_sf(data = dbisf,  size = 1.2, alpha = 0.6,colour = "dark blue") +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
  scale_colour_viridis_c() +
  theme_bw()+
  labs(title = paste0("Red Knot Delaware Bay locations (tag = 260808)"))+
 # guides(col= guide_legend(title= "Clusters"))+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

global_basic

```

<br>

We can measure the distance between each point (m) and plot these values to see if there are clear thresholds. This can help inform our choice of distance (m) in which a point is determined to be within the stopover "cluster" or an outlier. Previous work indicated a threshold of 2500m (2.5 km) distance buffer, however we also tested a broader definition of 5000m. In the figure below the horizontal lines represents the two thresholds 5000m and 2500m.

```{r, echo = FALSE, fig.width = 5, fig.cap= "Figure 2: Distance between points for tag 260808", fig.height=4, fig.align='centre'}
dbscan::kNNdistplot(data, k =  5)
abline(h = 5000, lty = 2)
abline(h = 2500, lty = 2)

```

<br>

We now run the density clustering algorithm. This requires two input parameters 1) minimum number of points for a cluster and 2) the distance threshold (eps) for points to be considered neighbors.

From a review of the figure 2 and literature we tested two distance thresholds; 1) 2500m and 2) 5000m. Figure 3 shows the result of the two different thresholds. Clusters listed as zero (0) are identified as outliers. We can see in Figure 3 that the smaller threshold distance (2500) results in identification of four clusters. This might be helpful in identifying sub locations within a broader stopover location. Alternatively when a wider threshold if used (5000m) the algorithm includes points in a single cluster, or marks them as outliers

```{r fig.height= 6, fig.width=8, echo =FALSE, fig.align='centre', fig.cap="Figure 3:Identified clusters using threshold values of 2500 (left) and 5000 (right). Colours represent identified clusters" }
global
```

<br>

Irrespective of the threshold chosen, we can filter out points identified as outliers to provide a data set of only stopover locations which forms the basis for our estimate of stopover locations.

<br> <br>

#### 2.3 Kernel Density Estimates (kde)

To better map the stopover areas we can use a kernel density estimate (kde) method to estimate the distribution of locations. Calculations for KDE require consideration of the mathematical method to describe the density function. A detailed description of the parameters can be found [here](http://www.spatialecology.com/gme/kde.htm).

KDEs are very sensitive to input parameters, specifically the bandwidth (h) which determines the [smoothing parameter](https://cran.r-project.org/web/packages/adehabitatHR/vignettes/adehabitatHR.pdf). H parameters can be estimated using three methods:

-   a reference bandwidth (h = σ × n \^−1/6), however this is generally an overestimate of the range, and is not suitable for multi-modal distributions.

-   Least Square Cross Validation (LSCV), which minimizes the difference in volume between the true UD and the estimates UD.

-   A subjective visual choice for the smoothing parameter, based on successive trials (Silverman, 1986; Wand & Jones 1995).

Other parameters include:

-   Kernel Type: The type of kernel is limited to Gaussian (bivariate normal), quadratic or normal. We used the bivariate normal model as default.

-   Grid/extent: These determines the area or extent over which the home range will be estimated. This is a mix of fine scale and time consuming processing and faster blocky resolution over a continuous surface. As a rule of thumb, Geospatial Modelling Environment program (GME) formally Hawth's tools suggest: take the square root of the x or y variance value (whichever is smaller) and divide by 5 or 10 (I usually round to the nearest big number - so 36.7 becomes 40). Before using this rule of thumb value calculate how many cells this will result in for the output (take the width and height of you input points, divide by the cell size, and multiply the resulting numbers together). If you get a value somewhere between 1-20 million, then you have a reasonable value.

<br>

We can review the impact of these parameters and different approaches for our Red Knot example tag to assess the impact on the projected stopover range. It is also possible to run KDE with set barriers and boundaries, which may be helpful for defining coastlines.

<br>

##### kde - method 1: h reference parameter

We estimate the stopover range using the "href" metrics. This is determined based on the spread of data. Once the range is calculated (**Utilization distribution of several Animals**) we can extract the vertices or polygons which define the percentage we wish to include. In the figure below we show the 95% threshold in yellow, 75% threshold is blue and 50% threshold is purple. The larger the percent selected the more inclusive the area mapped.

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 4: KDE estimates using href method. Polygons represnt the 95% threshold in yellow, 75% threshold is blue and 50% threshold is purple"}
## Generate kernal density ####

# #### 3.1 kde: h reference parameter
dbisfgeo <- dbisf |>  
  dplyr::filter(dbisf$cluster_2500 != 0) |> 
  dplyr::filter(eps == 2500) |> 
  dplyr::select(tag.id) |>
  as("Spatial")


# # define the parameters (h, kern, grid, extent) 
kde_href  <- kernelUD(dbisfgeo, h = "href", kern = c("bivnorm"), grid = 500, extent = 2)

# kde_href
# From this object (**Utilization distribution of several Animals**) we can extract the vertices or polygons which define the percentage we wish to include.
# 
# ```{r}
 ver95 <- getverticeshr(kde_href,95) # get vertices for home range
 ver95_sf <- st_as_sf(ver95)         # convert to sf object 
   
 ver75 <- getverticeshr(kde_href,75)
 ver75_sf <- st_as_sf(ver75 )
   
 ver50 <- getverticeshr(kde_href,50)
 ver50_sf<- st_as_sf(ver50)
  
# entire north America 
global_href <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  geom_sf(data = ver95_sf, alpha = 0.4, fill = "yellow")+
  geom_sf(data = ver75_sf, alpha = 0.4, fill = "dark blue")+
  geom_sf(data = ver50_sf, alpha = 0.4, fill = "purple")+
  #geom_sf(data = dbisf, , size = 1, alpha = 0.2,colour = "grey") +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
  theme_bw()+
  labs(title = paste0("KDE - h ref"))+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

global_href
 
```

<br>

##### kde - method 2: Least Squares Cross Validation (lscv) method.

The Least Square Cross Validation (LSCV) method, which minimizes the difference in volume between the true UD and the estimates UD, reflects a more restricted distribution. There is potential to refine the parameters to tweak these further.

```{r, echo = FALSE,warning=FALSE, message=FALSE, fig.cap="Figure 5: KDE estimates using least-square cross validation method."}
 
# #### 3.2 kde: Least Squares Cross Validation (lscv) method.
# 
kde_lscv  <- kernelUD(dbisfgeo, h = "LSCV", kern = c("bivnorm"), grid = 500, extent = 2)
 
ver95ls <- getverticeshr(kde_lscv,95) # get vertices for home range
ver95ls_sf <- st_as_sf(ver95ls) 
 
ver50ls <- getverticeshr(kde_lscv,50)
ver50ls_sf <- st_as_sf(ver50ls) 

# entire north America 
global_lscv <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  geom_sf(data = ver95ls_sf , alpha = 0.4, fill = "yellow")+
  #geom_sf(data = ver75_sf, alpha = 0.4, fill = "dark blue")+
  geom_sf(data = ver50ls_sf, alpha = 0.4, fill = "purple")+
  #geom_sf(data = dbisf, , size = 1, alpha = 0.2,colour = "grey") +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
  #scale_colour_viridis_c() +
  theme_bw()+
  labs(title = paste0("KDE - lscv"))+
 # guides(col= guide_legend(title= "Clusters"))+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

global_lscv

```

<br>

##### kde - method 3: variable smoothing parameters (h)

To test the sensitivity of the h value we can also test a bivariate normal kernel with a variety of smoothing parameters (h = 500, 1000, 3000). We can then interpret visually to determine size to use based on successive trials. This approach is supported by literature (Hemson et al. 2005; Calenge et al. 2011).

```{r, echo = FALSE, fig.align='center', fig.width=8, fig.height=5,warning=FALSE, message=FALSE, fig.cap="Figure 6: KDE estimates using various h smoothing parameters"}

kde_h1000  <- kernelUD(dbisfgeo, h = 1000, kern = c("bivnorm"), grid = 500,extent = 2)
kde_h500  <- kernelUD(dbisfgeo, h = 500, kern = c("bivnorm"), grid = 500,extent = 2)
kde_h3000  <- kernelUD(dbisfgeo, h = 3000, kern = c("bivnorm"), grid = 500,extent = 2)

# Lets extract the vertices and compare the outputs by building a plot.
# # kde - href = 1000
ver95_1000 <- getverticeshr(kde_h1000, 95) # get vertices for home range
ver95_1000_sf <- st_as_sf(ver95_1000) |>     mutate(h = 1000) # convert to sf object 

# # kde - href = 500
ver95_500 <- getverticeshr(kde_h500, 95) # get vertices for home range
ver95_500_sf <- st_as_sf(ver95_500)  |> 
   mutate(h = 500) # convert to sf object 
#   
# kde - href = 3000
ver95_3000 <- getverticeshr(kde_h3000, 95) # get vertices for home range
ver95_3000_sf <- st_as_sf(ver95_3000)  |> 
   mutate(h = 3000) # convert to sf object 
#   
# # bind all data together 
all_verts <- bind_rows(ver95_1000_sf,  ver95_500_sf,  ver95_3000_sf)

# entire north America 
global <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  geom_sf(data = all_verts, aes(colour = h), alpha = 0.1, colour = "dark blue") +
  #geom_sf(data = dbisf, size = 1.2, alpha = 0.2, colour = "dark blue") +
  #facet_wrap(~movement_dir)+
  # geom_point(ru, aes(x = lng, y = lat), size = 4) +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
  #coord_sf(xlim = c(-130, -40), ylim = c(10, 80), expand = FALSE)+
  #scale_colour_viridis_d() +
  #coord_sf() +
  labs(title = paste0("KDE - variable smoothing parameters"))+
  facet_wrap(vars(h)) +
  theme_bw()+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

global

```

<br>

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.height=4, fig.width=6}
##### Comparison of kde parameters
# ## Comparison of kde parameters
# 
# Lets rerun the home range by season and compare the methods for the KDE. 
dbisfgeo <- dbisf |>  
  dplyr::select(tag.id) |>
  as("Spatial")
 
 # href 
kde_href  <- kernelUD(dbisfgeo, h = "href", kern = c("bivnorm"), grid = 500, extent = 2)
 
 # custome h values 
 kde_h1000  <- kernelUD(dbisfgeo, h = 1000, kern = c("bivnorm"), grid = 500,extent = 2)
 
 # lscv 
 kde_lscv  <- kernelUD(dbisfgeo, h = "LSCV", kern = c("bivnorm"), grid = 500, extent = 2)
  
 
 # build function to get vertices 
 
get_verts <- function(in_kde, percent = 95, fieldname){

  ver <- getverticeshr(in_kde, percent) # get vertices for home range
  ver_sf <- st_as_sf(ver)  |>
    mutate(type = fieldname)

  return(ver_sf)
}

#dbisfgeo
# genertate vertices for each KDE

href <- get_verts(kde_href, percent = 95, fieldname = "href")
h1000 <- get_verts(kde_h1000, percent = 95, fieldname = "h1000")
kde_lscv <- get_verts(kde_lscv, percent = 95, fieldname = "lscv")
 
# bind all data together 
all_seasons <- bind_rows(href, h1000, kde_lscv)

# Geographic distributon of tags
world <- ne_countries(scale = "medium", returnclass = "sf")

Americas <- world %>% 
  dplyr::filter(region_un == "Americas")%>% 
  select(admin)

# entire north America 
global <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  geom_sf(data = all_seasons, aes(colour = id), alpha = 0.1, colour = "dark blue") +
  #geom_sf(data = dbisf, size = 1, alpha = 0.1, colour = "dark blue") +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
  facet_wrap(vars(type)) +
  theme_bw()+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

#global

```

<br>

### 4. Multiple individuals

We can apply these methods to multiple individuals to assess density of stopover locations within Delaware Bay. We can do this based on 1) stopover overlaps or 2) point density.

First, we subset individuals with had 10 or more locations within the area, we then use the dbscan method discussed above to isolate clusters or stopovers using the 2500m distance parameters. We then remove the outlier location points. Figure 8 below shows all gps/satellite tag locations within the Delaware Bay region over for 84 tags. The grey locations represent outliers and the blue locations are identified clusters, as generated per individual tag.

```{r, echo = FALSE, fig.align='center', fig.cap="Figure 7: GPS locations for 84 tags within the Delaware Bay region. Grey points are identified as outliers using density scan algorith per bird. Blue points are identified within the key stopovers", message=FALSE, warning=FALSE}

# read in del birds 
del_birds <- st_read(path("01_raw_data", "del_birds.gpkg"), quiet = TRUE)
db_sum <- del_birds |> 
  st_drop_geometry() |> 
  group_by(tag.id) |> 
  summarise(n = n()) |> 
  arrange(n)

# filter where less than 10 points within aoi
db_tagids <- db_sum |> 
  filter(n > 10) |> 
  pull(tag.id)


#####################################
## 1: Use db scan to remove outliers based on month and tag id. 

db <- del_birds |> filter(tag.id %in% db_tagids) 
db <- st_transform(db, crs = 4087)
db <- cbind(db, st_coordinates(db))

#plot(db$year, db$month)
#hist(db$month)

db <- db |>
  st_drop_geometry() |> 
  select(tag.id,year, month, day, X, Y)

dbsf <- st_as_sf(db, coords = c("X", "Y"), crs = 4087)

# make list of tags by month 
#tagls <- unique(db$tag.id)
#taglsm <- db |> select(tag.id, month, day) |> unique()
#taglsm <- taglsm |> mutate(id = seq(1:length(taglsm$tag.id)))


all_pts <- st_read(path("01_raw_data", "Del_birds_month_yr_pts.gpkg"), quiet = TRUE)
# pts with remove the outlier points 

# # # Geographic distributon of tags
world <- ne_countries(scale = "medium", returnclass = "sf")
Americas <- world %>%
   dplyr::filter(region_un == "Americas")%>%
   select(admin)
 
# entire north America
global <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  geom_sf(data = dbsf, size = 1.2, alpha = 0.4, col = "darkgrey") +
  geom_sf(data = all_pts, size = 1.2, alpha = 0.2, col = "blue") +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-76, -73.5), ylim = c(38, 40.5), expand = FALSE)+
  scale_colour_viridis_c() +
  theme_bw()+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())
global

```

<br>

#### 4.1 Stopover polygon density

As described above can generate a home range or stopover polygon for each bird using the KDE method. For this summary we used a h-ref smoothing function and generated a home range kernel for each tag and month combination. All years were combined but these could be split out as needed.

```{r, echo = FALSE, fig.align='center', fig.cap="Figure 8: GPS points and estimates stopover ranges per month for Delaware Bay", fig.width = 8, fig.height=8}
## Summarise the dataset by tag id and no of months 

all_polys <- st_read(path("01_raw_data", "Del_birds_month_yr_polygons.gpkg"), quiet = TRUE)

all_pts <- st_read(path("01_raw_data", "Del_birds_month_yr_pts.gpkg"), quiet = TRUE)

# summary_db <- all_polys |> 
#   st_drop_geometry() |> 
#   group_by(id) |> 
#   summarise(n_months = n_distinct(month),
#             n_years = n_distinct(year)) 
# 

## note we can filter this down to 50% percentage 

all_poly_50 <- all_polys |> filter(th == 50)

# ggplot 

global_50 <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  #geom_sf(data = dbsf,  size = 1, alpha = 0.2,colour = "blue") +
  geom_sf(data = all_pts,  size = 1, alpha = 0.2,colour = "blue") +
  #geom_sf(data = all_poly_50, linewidth = 0.5, alpha = 0.4, fill = "red")+
  geom_sf(data = all_poly_50, linewidth = 0.1,  alpha = 0.2,fill = "red", color = "red")+
  xlab("Longitude") + ylab("Latitude") +
  #coord_sf(xlim = c(-75, -74.35), ylim = c(38.8, 39.4), expand = FALSE)+
  coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
  theme_bw()+
  facet_wrap(~month)+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

global_50


```

<br>

In figure 8 above, we can see the refined point locations (blue) along with a home range (red polygons) for the 84 individuals who were tracked through Delaware Bay. We plot each month separately to view how stopover locations change with season.

#### 4.2 Stopover point density.

Alternately we can review the density of stopover use based on the points rather than estimated home range (polygons). To ensure this was a standardized metrics, I selected only one location per day (rather than the full data set used above). We can then generate density metrics over all months and years (Figure 9) or by month (Figure 10).

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.cap="Figure 9: Point Density estimates (all months combined)"}

###################################################
## Part 3: Calculate a density plots using points. 

##################################################
# # filter so that all points are the same density 
library(viridis)
library(viridisLite)
library(ggpointdensity)
library(ggspatial) # For base maps

all_pts <- st_read(path("01_raw_data", "Del_birds_month_yr_pts.gpkg"), quiet = TRUE)

all_pts<- cbind(all_pts, st_coordinates(all_pts))

all_pts_density <- all_pts|> group_by(tag.id, year, month, day) |>
  slice_head(n = 1) |> 
  ungroup() 

ggplot(data = all_pts, mapping = aes(x = X, y = Y)) +
  geom_pointdensity() +
  scale_color_viridis()+
  theme_bw()
```

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.cap="Figure 10: Point Density estimates by month"}

ggplot(data = all_pts, mapping = aes(x = X, y = Y)) +
  geom_pointdensity() +
  #annotation_map_tile(type = "osm") +
  scale_color_viridis()+
  facet_wrap( ~ month ) 
# 
# 
# ggplot(data = all_pts, mapping = aes(x = X, y = Y)) +
#   geom_bin2d()+
#   #annotation_map_tile(type = "osm") +
#   scale_color_viridis()+
#   facet_wrap( ~ month ) 


# 
# ## Testing map 1 
# 
# ggplot(all_pts) +
#   stat_density_2d(aes(x = st_coordinates(all_pts)[,1],
#                       y = st_coordinates(all_pts)[,2]),
#                   geom = "polygon") +
#   scale_fill_viridis_c() +
#   labs(title = "Spatial Point Density", x = "Longitude", y = "Latitude") +
#   theme_minimal()

# Testing map 2
# 
# global_href <- ggplot(data = Americas) +
#   geom_sf(color = "grey") +
#   #geom_sf(data = dbsf,  size = 1, alpha = 0.2,colour = "blue") +
#   #geom_sf(data = all_pts,  size = 1, alpha = 0.2,colour = "lightblue") +
#   geom_pointdensity(all_pts, mapping = aes(x = X, y = Y)) +
#   
#   #geom_sf(data = all_poly_50,  alpha = 0.4,fill = "red")+
#   xlab("Longitude") + ylab("Latitude") +
#   #coord_sf(xlim = c(-75, -74.35), ylim = c(38.8, 39.4), expand = FALSE)+
#   coord_sf(xlim = c(-76, -74), ylim = c(38, 40), expand = FALSE)+
#   theme_bw()+
#   facet_wrap(~month)+
#   theme(axis.text.x=element_blank(),
#         axis.text.y=element_blank())
# 
# global_href

```

We can see these are fairly similar to the kde estimates.

<br> <br>

## 5. Questions for discussions

To identify stopover locations we have to decide on which approach best fits the Red Knot stopover behavior and ecology.

-   Distance threshold for stopover locations. Figure 3 show the different clusters identified by using two different distance threshold values. I would suggest sticking with finer scale (2500), as this might be helpful in differentiating sub clusters in the future as needed.

-   Kde smoothing parameter. Based on the testing above I would suggest using a wider smoothing function (h value), then generating the % thresholds as shown in Figure 4. Depending on how tightly you want to refine the stopover locations we can calculate the stopover location and then use a standard threshold to narrow down the area used. An example is show in Figure 11 below. In this figure we use a href approach with various % thresholds (50%, 65%, 75%, 85%,and 95% ) to illustrate the variability.

```{r, echo = FALSE, message = FALSE, fig.align='center', fig.width=8, fig.height=6, warning=FALSE, fig.cap="Figure 11: KDE estimates using href smoothing and various percent thresholds" }
dbisfgeosf <- st_as_sf(dbisfgeo)

# # define the parameters (h, kern, grid, extent) 
kde_href  <- kernelUD(dbisfgeo, h = "href", kern = c("bivnorm"), grid = 500, extent = 2)

 ver95 <- getverticeshr(kde_href,95) # get vertices for home range
 ver95_sf <- st_as_sf(ver95) |> 
   mutate(th = 95)        # convert to sf object 
 
 ver85 <- getverticeshr(kde_href,85)
 ver85_sf <- st_as_sf(ver85 )|> 
   mutate(th = 85)
   
 ver75 <- getverticeshr(kde_href,75)
 ver75_sf <- st_as_sf(ver75 )|> 
   mutate(th = 75)
   
 ver65 <- getverticeshr(kde_href,65)
 ver65_sf <- st_as_sf(ver65 )|> 
   mutate(th = 65)
 
 ver50 <- getverticeshr(kde_href,50)
 ver50_sf<- st_as_sf(ver50) |> 
   mutate(th = 50)
  
 allvers<- bind_rows( ver95_sf,ver85_sf, ver75_sf , ver65_sf, ver50_sf)
 
 # entire north America 
global_href <- ggplot(data = Americas) +
  geom_sf(color = "grey") +
  geom_sf(data = dbisfgeosf, , size = 1, alpha = 0.2,colour = "blue") +
  geom_sf(data = allvers, linewidth = 0.5, alpha = 0.4, fill = "red")+
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(-75.2, -74.35), ylim = c(38.8, 39.4), expand = FALSE)+
  theme_bw()+
  facet_wrap(~th)+
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

global_href
 
```

<br>

-   Multiple species analysis-How should the home range estimates be calculated temporally? i.e. months or potential north/south ward direction? This could be tricky for birds who winter in South East USA.

-   Preference for polygon (kde) vs point density. I think using kde is a more robust option as generated per tag and time period as this accounts for individual bird movement, not captured using a point density metrics directly.

-   Currently this is limited to rufa subpopulations? Is this correct?

<br>

## 6. Next Steps

-   Decide on parameters for dbscan - i.e. 2500m or 5000m
-   Use full data set to identify clusters for all birds.

<br>

<br>

### References

-   Packages: Estimate Kernel home range Utilization Distribution Using adehabitatHR [(Calenge et al. 2011)](https://cran.r-project.org/web/packages/adehabitatHR/vignettes/adehabitatHR.pdf)

-   [KDE](%22http://www.spatialecology.com/gme/kde.htm%22)

-   https://www.ckwri.tamuk.edu/sites/default/files/publication/pdfs/2017/leonard_analyzing_wildlife_telemetry_data_in_r.pdf

-   Seaman, D. E., Millspaugh, J. J., Kernohan, B. J., Brundige, G. C., Raedeke, K. J., & Gitzen, R. A. (1999). Effects of sample #size on kernel home range estimates. The journal of wildlife management, 739-747.

-   Kernohan, B. J., R. A. Gitzen, and J. J. Millspaugh. 2001. Analysis of animal space use and movements. Pages 125--166 in J. J. #Millspaugh and J. M. Marzluff, editors. Radio tracking and animal populations. Academic Press, San Diego, CA, USA

-   Hemson, G., Johnson, P., South, A., Kenward, R., Ripley, R., & MACDONALD, D. (2005). Are kernels the mustard? Data from global positioning system (GPS) collars suggests problems for kernel home‐range analyses with least‐squares cross‐validation. Journal of Animal Ecology, 74(3), 455-463
